The hedging hyperprior for the normal distribution: known mean, unknown precision

### 2023-03-28 text/23/normal-hedging-hyperprior-part-2.Rmd
More formula updates on hedging hyperprior

Consider a Bayesian analysis of an outcome variable that has a normal distribution.

For this derivation, it is easier to write the density in terms of the precision

$f(X; \mu, \theta) = \sqrt{\frac{\theta}{2 \pi}}e^{-\frac{1}{2}\theta (X-\mu)^2}$

where

$\theta=\frac{1}{\sigma*2}$

Assume that the mean of the normal distribution, $\mu$, is known and equal to $\mu_F$. An earlier case with precision known and the mean  unknown appears on a different web page. A more general case, with both the mean and the precision of the normal distribution unknown is also described on a separate web page. For simplicity, we will drop any multiplicative constants not involving X or $\theta$

$f(X; \theta) \propto \sqrt{\theta}e^{-\frac{1}{2}\theta(X-\mu_F)^2}$

@@

$g(\theta; \alpha_0, \beta_0) = \frac{\beta_0^{\alpha_0}}{\Gamma(\alpha_0)} \theta^{\alpha_0-1} e^{-\beta_0 \theta}$

@@

$\beta_0 = n_0 \theta_0$

and remove any multiplicative constants with respect to $\theta$ to get

$g(\theta; n_0, \theta_0) \propto \theta^{n_0-1} e^{-n_0 \theta_0 \theta}$

@@

$L(\mathbf{X} | \mu) \propto \theta^{n/2}e^{- \frac{1}{2} \theta\sum(X_i-\mu_0)^2}$

@@

$h(\theta | \mathbf{X}) \propto \theta^{n_0-1} e^{-n_0 \theta / \theta0} \theta^{n/2} e^{- \frac{1}{2} \theta\sum(X_i-\mu_0)^2}$

@@

$h(\theta | \mathbf{X}) \propto \theta^{n/2 + n_0-1} e^{-(n_0 \theta0 + \frac{1}{2}\sum(X_i-\mu_0)^2)}$

@@

$\beta_1 = n_0 theta_0 + \frac{1}{2}\sum(X_i-\mu_0)^2)$

This can also be shown to relate to a t-distribution

$\frac{\Gamma \left(\frac{\nu+1}{2} \right)} {\sqrt{\nu\pi}\,\Gamma \left(\frac{\nu}{2} \right)} \left(1+\frac{x^2}{\nu} \right)^{-\frac{\nu+1}{2}}$

@@

n0 <- 4
theta0 <- 0.5

alpha0 <- n0
beta0 <- n0*theta0

@@

x <- muF+ round(rnorm(n), 1)/theta0

@@

f <- function(x, mu, theta) {
  sqrt(theta/2*pi)*exp(-0.5*theta*(x-mu)^2)
}

@@

Let's illustrate this with an informative prior with $n_0$ = `r n0` and $\theta_0$ = `r theta0`.

@@

This corresponds to a fairly informative prior distribution. This distribution places $\theta$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.

@@

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r theta0` and it excludes `r ss/(n-1)`, the mean of the data.

@@

mean $\theta_0$ and prior precision

@@

$\beta = \alpha \theta_0 $

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

gstar <- function(theta, lambda) {g(theta, 1+lambda*(alpha0-1), (1+lambda*(alpha0-1))*beta0)}
density <- outer(theta_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

This is the joint prior distribution of $\theta$ and $\lambda$.

It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\theta$ is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r, eval=FALSE}
density <- outer(theta_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The likelihood function is constant across all values of $\theta$, so its surface looks like the following.

```{r, eval=FALSE}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

fstar <- function(mu, lambda) {f(xbar, mu, theta0)^n}
likelihood <- outer(mu_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

hstar <- function(mu, lambda) {g(mu, mu0, lambda*(theta0-thetaF) + thetaF)*f(xbar, mu, thetaF)^n}
density <- outer(mu_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r, eval=FALSE}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  mu_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(50, 75, 90)/100,
  xlab="theta",
  ylab="lambda")
```

You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r theta0`) is largely ignored in favor of the mean of the data (`r "??"`).

>>>>

Consider a Bayesian analysis of an outcome variable that has a normal distribution with known mean and unknown precision. Assume that both the mean and the precision of the normal distribution are unknown. Other cases with [mean unknown and precision known][sim1] and with [both mean and precision unknown][sim3] appear on a different web pages.

[sim1]: http://new.pmean.com/normal-hedging-hyperprior-part1
[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part3

Assume that the mean of the normal distribution, $\mu$, is known and equal to $\mu_F$. An earlier case with precision known and the mean  unknown appears on a different web page. A more general case, with both the mean and the precision of the normal distribution unknown is also described on a separate web page. For simplicity, we will drop any multiplicative constants not involving X or $\tau$

$f(X; \tau) \propto
  \sqrt{\tau}
  e^{-\frac{1}{2}\tau(X-\mu_F)^2}$

@@

$g(\tau; \alpha_0, \beta_0) =
  \frac
    {\beta_0^{\alpha_0}}
    {\Gamma(\alpha_0)}
  \tau^{\alpha_0-1}
  e^{-\beta_0 \tau}$

@@

$\beta_0=n_0\tau_0$

and remove any multiplicative constants with respect to $\tau$ to get

$g(\tau; n_0, \tau_0) \propto \tau^{n_0-1} e^{-n_0 \tau_0 \tau}$

@@

$L(\mathbf{X} | \mu) \propto
  \tau^{n/2}
  e^{- \frac{1}{2} \tau\sum(X_i-\mu_0)^2}$

@@

$h(\tau | \mathbf{X}) \propto
  \tau^{n_0-1}
  e^{-n_0 \tau / \tau0}
  \tau^{n/2}
  e^{- \frac{1}{2} \tau\sum(X_i-\mu_0)^2}$

@@

$h(\tau | \mathbf{X}) \propto
  \tau^{n/2 + n_0-1}
  e^{-(n_0 \tau0 + \frac{1}{2}\sum(X_i-\mu_0)^2)}$

@@

$\beta_1 = n_0 \tau_0 + \frac{1}{2}\sum(X_i-\mu_0)^2)$

### A simple numerical example

@@

n0 <- 4
tau0 <- 0.5

alpha0 <- n0
beta0 <- n0*tau0

@@

x <- muF+ round(rnorm(n), 1)/tau0

@@

f <- function(x, mu, tau) {
  sqrt(tau/2*pi)*exp(-0.5*tau*(x-mu)^2)
}

@@

Let's illustrate this with an informative prior with $n_0$ = `r n0` and $\tau_0$ = `r tau0`.

@@

This corresponds to a fairly informative prior distribution. This distribution places $\tau$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.

@@

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\tau$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r tau0` and it excludes `r ss/(n-1)`, the mean of the data.

@@

mean $\tau_0$ and prior precision

@@

$\beta = \alpha \tau_0 $

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
tau_grid <- seq(0, x_max, length=60)

gstar <- function(tau, lambda) {g(tau, 1+lambda*(alpha0-1), (1+lambda*(alpha0-1))*beta0)}
density <- outer(tau_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  tau_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="tau",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

This is the joint prior distribution of $\tau$ and $\lambda$.

It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\tau$ is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r, eval=FALSE}
density <- outer(tau_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="tau",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The likelihood function is constant across all values of $\tau$, so its surface looks like the following.

```{r, eval=FALSE}
lambda_grid <- seq(0, 1, length=60)
tau_grid <- seq(0, x_max, length=60)

fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
likelihood <- outer(mu_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
density <- outer(mu_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r, eval=FALSE}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  mu_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(50, 75, 90)/100,
  xlab="tau",
  ylab="lambda")
```

You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r tau0`) is largely ignored in favor of the mean of the data (`r "??"`).
